Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-26 methods.tex
minor edits to methods.tex incl regime -> segment
moved seizure detection citation to concl

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of an
rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization changes quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \svd. Each point is the average IPC
      during a 100,000 instruction period.  The colors refer to different
``regimes'' of the signal, as discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
cycles), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track and understand.

This wide range of behaviors provides a unique advantage, for the
purposes of this paper, in that a number of different generating
processes---with a wide range of complexities---are at work at
different points in a single time series.  The complexity of \col and
\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
far more consistent over time---probably the result of a single
generating process.  \svd, in contrast, has multiple regimes, each the
result of a different generating process.  To take advantage of this,
we split the signal into six different segments, thereby obtaining an
array of examples for the analyses in the following sections.  For
notational convenience, we refer to these 90 time series\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
segments of \svd, ordered from left to right.  These segments, which
were determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure\footnote{It should be noted that a version of the complexity
  metric discussed in Section~\ref{sec:meaComplex} has been used to
  detect seizures in brainwave data \cite{cao2004det}, which can be
  viewed as regime detection.  The purpose of this paper is \emph{not}
  to rigorously explore regime detection but to explore quantifying
  complexity of a time series.}.  Thus, it may the case that more than
one generating process may be at work in each of our segments.  This
is a factor in the discussion of Section~\ref{sec:results}.

>>>>

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of an
rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization changes quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \svd. Each point is the average IPC
      during a 100,000 instruction period.  The colors refer to
      different {\sl segments} of the signal, as discussed in the
      text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
cycles), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track and understand.

This wide range of behaviors provides a unique advantage, for the
purposes of this paper, in that a number of different generating
processes---with a wide range of complexities---are at work at
different points in a single time series.  The complexity of \col and
\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
far more consistent over time---probably the result of a single
generating process.  \svd, in contrast, has multiple regimes, each the
result of a different generating process.  To take advantage of this,
we split the signal into six different segments, thereby obtaining an
array of examples for the analyses in the following sections.  For
notational convenience, we refer to these 90 time series\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
segments of \svd, ordered from left to right.  These segments, which
were determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may the case that more than one generating
process may be at work in each of our segments.  This is a factor in
the discussion of Section~\ref{sec:results}.
